Future challenges for Research Software Engineering

### 2025-05-14 future_challenges_for_RSEng.tex
Proof read up to end of section 2.6.

Research Software Engineering is currently an emerging field,
with an increasing integration into various aspects of research.
In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living.
In the following sections, we consider the evolution of the RSE definition,
the development of RSE education,
the growth of the RSE communities,
the impact of automation and generative artificial intelligence in RSE work,
the emerging ethical considerations,
and the transformations expected in the research organisations,
and if there will be other in academia covered by new RTPs (\emph{Research Technical Professional}s).
We conclude the paper with a summary and with our wishes and recommendations.

\section{Aspects of this future}
%\subsection{What do we want to have achieved by 2035}
Looking ahead over a 10-year timeframe, we hope that the RSE community will have
contributed to make research software and its outputs
more sustainable, robust and maintainable.
We also hope, that Research Software becomes an accepted research artifact on equal footing as data and publications.
We see this happening in a number of ways.
However, we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
With development of software tools and scripted workflows becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

@@

So where do RSEs fit into this picture? Even today we expect to see RSEs providing technical skills and input
to research activities in a range of different areas, such as
High-quality, robust architecture and design for research software,
Seamless integration with research and data infrastructures,
Development of software for novel and emerging architectures,
Application of specialist numerical and statistical methods within software,
Optimisation of codes for use on high-performance computing infrastructure,
and Green computing - efficient implementations of algorithms/methods and use of hardware

It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}
and it is even more hopeless to structure the aspects that will become relavant for the RSE community in the future.
Nevertheless, we decided to consider
\begin{itemize}
\item The evolution of the term RSE,
\item Development of the RSE community,
\item What the impact of AI to the work of RSEs will look like,
\item the impact of Complexity and Usability in Research Software
\item Ethics and social considerations,
\item How education is reshaped by the digital transformation,
\item Changes to the Academic Institutions,
\item Impact of alternate career paths for RSEs and other RTPs.
\end{itemize}

\begin{story}{Introduction}{intro}
In order to give a more explicit setting, imagine a world ten years from now.
Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
But the world has moved on.
When they look back, what will they see as problems which are now solved?
Which tasks are persisting, where they are now happy that a younger generation is now taking up the baton and carries forward their work.
What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.\\
In order to bring this future to life these two personas will accompany us and we will shed light on certain aspects
of their future in respective small story boxes.
 \end{story}

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{}{definition-evolution-rse}
The term RSE is established in the minds of university boards, as a valuable member in the academic framework.
RSEs are developed using specialized master's program, and their support is valued by researchers. We still teach researchers basic software skills, but the content of this lecture has changed drastically.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalisation of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
In this discussions the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?)
(GC: Higher-level and more intuitive workflows that require less training).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

What will \emph{not} change over that timeframe is the RSE's ability to remain
flexible and embrace change. Keep an eye open and pick up tools and skills
as needed as we go along.
Communication and collaboration skills will remain essential to engage
with domain scientists and help them find technical solutions.\\

@@

\subsection{Development of the RSE community}
\begin{whatis}{}{development-rse-community}
National RSE communities are increasingly collaborating with local partners, fostering stronger networks across countries. Regular international conferences further amplify these connections, promoting the global exchange of ideas. The introduction of RSE Master's programs, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile. This growing recognition has led to a surge in membership within national societies, reflecting the increasing importance of RSEng in the research landscape.
\end{whatis}
In recent years, we saw the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation%
\footnote{MM: do we want to decide for UK or US English? Or keep it mixed? Not being a native English speaker I don't know how it reads.}
of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSEs to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
become available at various universities.
With increased numbers the visibility of the RSE community also to industry will become higher.
\begin{story}{Development of RSE community}{RSEcommunity}
Kay stood proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community. Having moved to South Africa, she spoke passionately about the unique challenges researchers in the Global South face—limited access to high-performance computing, unreliable funding, and the persistent digital divide. Her keynote highlighted how local innovation thrives despite constraints, often out of necessity rather than choice. The audience, a diverse mix of global experts, listened intently as she called for equitable collaborations, not just inclusion as an afterthought.
\end{story}

\subsection{Impact of AI and automation}
\begin{whatis}{}{impact-ai-automation}
RSEs are educated about AI models, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modeling and predictive tasks.
For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
Developers are shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
The integration of AI is democratizing access to advanced technologies and analytics, enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster, more efficient, and more innovative than ever before.
With the general availability of AI the interaction with researchers will change at a fundamental level.
Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
It can be foreseen, that AI drives this low-key interaction to a highly specialized level of knowledge,
since AI copilots are available 24/7.
Since AI systems are so democratically available and will be used as black boxes,
the interpretation of their outputs will be far more imortant.
Currently we are expanding and documenting our work in the open and making our code available,
therefore there will be more structured material to train GenAI on.
However, as people get quicker answers from GenAI, the community resources (\eg Stack Overflow, Wikipedia)
get poorer.

\begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
Kim develops advanced scientific software in tandem with an AI agent that codes, debugs, and simulates experiments alongside her. The AI proactively suggests optimisations, generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs. With this partnership, Kim accelerates her research output while maintaining full control over scientific direction and ethical oversight.
\end{story}

\subsection{Impact of Complexity and Usability in Research Software}
\begin{whatis}{}{abstraction}
With a palette of readily reusable software components,
and with their skills in managing hierarchical systems of complex software components with clear interfaces,
RSEs can now build systems of scale and capability not previously possible.
\end{whatis}

@@

We acknowledge but would also like to challenge the observations of the participants.
We believe that ``we are standing on the shoulders of giants'':
By reducing the complexity of individual software components,
we are now able to develop systems of complexity and capability not previously possible.
This divide-and-conquer approach allows the scientific community to address questions not previously in grasp.
When it comes to teaching, it is true that students need to learn how to handle complexity,
by being able to develop deep understanding of individual components.
However, it is also important to learn how to manage complexity
by building hierarchical systems of well-encapsulated, reusable components.
It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
as well as RSEs that can manage complex systems of almost black-box components,
and the future of RSE education should aim to prepare students for both roles.

@@

\subsection{Ethics and social considerations}
\begin{whatis}{}{ethics}
The principles of Good Scientific Practice \cite{dfg_gsp} have been a cornerstone in the conduct of research.
But their relevance hinges on constant updates in light of how research is performed.
By 2035 we want to have achieved accepted rules on the use of AI in research tasks and software development.
Open Science and Open Research are well-established in academia, and supported by respective Open Research Software
strategies of the political actors. We also hope, that Data Awareness and Data Protection are also still values
that are protected.
\end{whatis}

@@

% institutional bonds and ethics
Sometimes these values are in direct conflict with the aims of their institution. For example, their PI might insist on quickly developing some software without adhering to current best practices.
A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
This can then form the basis for institutional policies and guidelines.

% society implications
The technology people working as RSE might have, well directly or indirectly, ethical implications in areas such as social engineering [TODO find source], climate [TODO find source] and more.
Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
Like engineers, RSEs take the responsibility for the software they develop.
RSEs need to be aware of these issues and develop their own ``red lines''.

@@

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay have started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is working in a project that develops
software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
at the university, and increasingly struggles to find compelling Open Source alternatives that lets them keep their data
to software that is backed by large corporate entities. So far she still has the support of the university board, since it values
strategic autonomy.
Her son Kyle on the other hand started out with a degree in Physics and then took one of the newly created RSE Master's courses,
but quickly decided that work in academia with its time limited contracts is not attractive.
Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
and therefore have deprioritized publicly funded research. Kyle decided to go to a defense company since they offer a
more secure employment in this polarized world.
\end{story}

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{}{academic}
We expect research software to be a central part of scientific practice on the same level as data and publications.
This will be supported by the first generations of RSE-Master graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, that are known by 2035 in university boards, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which have them, we want to see that researchers naturally accept them as part of their used services.
Additionally researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks. This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years such as
the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
But we could imagine new RSE specialisations to emerge in the fields
of quantum computing, such as the Quantum-RSE, and in federated online platforms for research.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

On the other hand the existence of people with a certain specilised skillset opens up possiblities for pooling.
Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernisation, cleanup, and modularisation,
which can pay off in faster and larger development in the future.
This is an opportunity that would not be possible by short-term funded solo RSEs.

\emph{Which challenges will still persist}\\
The workshop participants expect that there will still be budgeting and administrative hurdles to creating
permanent and well-funded RSE teams.
Here it can only be strongly hoped, that, with the recognized importance of research software, state-actors
support the development with a reliable stream of funding.
This would also require a change of culture in research about the acceptance of research software.\\
While we hope that the barrier to researchers diminishes, the workshop participants
still expect some work to do. We hope that the compartmentalisation that occurs will help researchers reprioritize their workload.
An upcoming challenge will be that, with the emergence of more dedicated RSE units,
networking between them will come into the foreground, because there should be a strong need to share best practices
among institutions with successful RSE Teams.
On a more organisational level, we expect the sharing of models on how how to organise diiferent RSEs, to fit the different organisational constraints.
After that Organisations will have to come up with models for hiring people.
While a standardized RSE Master's makes it easier, it will be interesting to see if
RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE master's degree ensures a hiring process on a more standardised level.

\begin{story}{Changes to the Universities}{changes}
Kim found work years ago at one of the first RSE units in Arcadia. After a couple of years he moved to the university of
Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
Kim improves upon the model set forth by the University of Arcadia based on valuable experience. Since the university
of Avalon has to follow different rules not everything can be straightforwardly adapted. Life would be so much easier,
if at least university regulations would be homogenized, no?
\end{story}

\emph{New challenges}
If we follow the ideas laid out in \cite{Kempf2025-draft}, then in addition to university wide RSE units,
there will be embedded RSEs in the departments. Shaping and developing these network bonds will be a task for the future.
While networking embedded RSEs and RSE units is essential for building robust, sustainable research software infrastructure, but the question arises whether every group truly needs to host and manage its own infrastructure.
When each team or institution hosts its own systems, this often leads to siloed environments where resources, expertise, and data are isolated, limiting collaboration and creating inefficiencies. Such silos can result in duplicated efforts, higher costs, and challenges in maintaining data integrity and governance, as each group independently manages similar infrastructure and workflows. While local control offers flexibility, it also increases complexity and makes it harder to share innovations or best practices across the wider research community. To overcome these challenges, federated infrastructure models are needed, enabling shared access, interoperability, and coordinated development while still allowing for some local autonomy \cite{Jelinek2021}.
Federated approaches break down silos, promote knowledge exchange, and ensure that infrastructure investments benefit the broader ecosystem rather than just isolated groups.

>>>>

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and advance the conversation about how Research Software Engineering should, or could, develop.
In the following sections, we consider a number of areas:
the evolution of the RSE definition;
the development of RSE education;
the growth of the RSE communities;
the impact of automation and generative artificial intelligence in RSE work;
the emerging ethical considerations;
the transformations expected in research organisations;
and, finally, if there will be other areas in academia covered by new groups of RTPs (\emph{Research Technical Professional}s).
We conclude the paper with a summary and with our wishes and recommendations.

\section{Aspects of this future}
%\subsection{What do we want to have achieved by 2035}
Looking ahead over a 10-year timeframe, we hope that the RSE community will have
contributed to making research software and its outputs
more sustainable, robust and maintainable.
We also hope that Research Software becomes an accepted research artifact on an equal footing with data and publications.
We see this happening in a number of ways,
particularly through movements such as DORA~\cite{DORA} and CoARA~\cite{COARA} which are advocating for a change in the way that research is assessed and outputs recognised.
However, we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
With development of software tools and scripted workflows becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

@@

So where do RSEs fit into this picture? Even today we expect to see RSEs providing technical skills and input
to research activities in a range of different areas, these include:
high-quality, robust architecture and design for research software;
seamless integration with research and data infrastructures;
development of software for novel and emerging architectures;
application of specialist numerical and statistical methods within software;
optimisation of codes for use on high-performance computing infrastructure;
and green computing - efficient implementations of algorithms/methods and use of hardware

It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}
and it is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
Nevertheless, we decided to consider:
\begin{itemize}
\item The evolution of the term RSE
\item Development of the RSE community
\item What the impact of AI will be to the work of RSEs
\item The impact of complexity and usability in research software
\item Ethics and social considerations
\item How education is reshaped by the digital transformation
\item Changes to academic institutions
\item Impact of alternate career paths for RSEs and other RTPs
\end{itemize}

\begin{story}{Introduction}{intro}
In order to give a more concrete setting, imagine a world ten years from now.
Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024},
our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
But the world has moved on.
When they look back, what will they see as problems which are now solved?
Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
What issues have newly emerged, where they are now just happy to say: “Oh well… my successors will take care of that”.
In order to bring this future to life, these two personas will accompany us through the rest of this paper and we will shed light on aspects
of their future in respective small story boxes.
 \end{story}

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{}{definition-evolution-rse}
The term RSE is established in the minds of university boards and leadership, as a valuable member in the academic framework.
New RSEs are trained through a specialized master's program, and their support is valued by researchers. We still teach researchers basic software skills, but the content of this lecture has changed drastically.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalisation of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
In this discussions the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?)
(GC: Higher-level and more intuitive workflows that require less training).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

What will \emph{not} change over that timeframe is the RSE's ability to remain
flexible and embrace change, and keep an eye on the wider landscape,
picking up new tools and skills as needed.
Communication and collaboration skills will remain essential to engage
with domain scientists and help them find technical solutions.\\

@@

\subsection{Development of the RSE community}
\begin{whatis}{}{development-rse-community}
National RSE communities are increasingly collaborating with local partners, fostering stronger networks across countries. Regular international conferences further amplify these connections, promoting the global exchange of ideas. The introduction of RSE Master's programs, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile. This growing recognition has led to a surge in membership within national societies, reflecting the increasing importance of RSEng in the research landscape.
\end{whatis}
In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation%
\footnote{MM: do we want to decide for UK or US English? Or keep it mixed? Not being a native English speaker I don't know how it reads.}
of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSEs to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
become available at various universities.
With increased numbers, the visibility of the RSE community to industry will be enhanced.
\begin{story}{Development of RSE community}{RSEcommunity}
Kay stood proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community. Having moved to South Africa, she spoke passionately about the unique challenges researchers in the Global South face—limited access to high-performance computing, unreliable funding, and the persistent digital divide. Her keynote highlighted how local innovation thrives despite constraints, often out of necessity rather than choice. The audience, a diverse mix of global experts, listened intently as she called for equitable collaborations, not just inclusion as an afterthought.
\end{story}

\subsection{Impact of AI and automation}
\begin{whatis}{}{impact-ai-automation}
RSEs are educated about AI models, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modeling and predictive tasks.
For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
Developers are shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
The integration of AI is democratizing access to advanced technologies and analytics, enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster, more efficient, and more innovative than ever before.
With the general availability of AI the interaction with researchers will change at a fundamental level.
Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
it can be foreseen that AI will drive this low-key interaction to a highly specialized level of knowledge,
since AI copilots are available 24/7.
Since AI systems are so democratically available and will be used as black boxes,
the interpretation of their outputs will be far more imortant.
Currently we are expanding and documenting our work in the open and making our code available,
therefore there will be more structured material to train GenAI on.
However, as people get quicker answers from GenAI, the community resources (\eg Stack Overflow, Wikipedia)
get poorer.

\begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
Kim develops advanced scientific software in tandem with an AI agent that codes, debugs, and simulates experiments alongside him. The AI proactively suggests optimisations, generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs. With this partnership, Kim accelerates his research output while maintaining full control over scientific direction and ethical oversight.
\end{story}

\subsection{Impact of Complexity and Usability in Research Software}
\begin{whatis}{}{abstraction}
With a palette of readily reusable software components,
and with their skills in managing hierarchical systems of complex software components with clear interfaces,
RSEs can now build reliable, maintainable systems of a scale and capability not previously possible.
\end{whatis}

@@

We acknowledge but would also like to challenge the observations of the participants.
We believe that ``we are standing on the shoulders of giants'':
By reducing the complexity of individual software components,
we are now able to develop systems of complexity and capability not previously possible.
This divide-and-conquer approach allows the scientific community to address questions not previously within their grasp.
When it comes to teaching, it is true that students need to learn how to handle complexity,
by being able to develop deep understanding of individual components.
However, it is also important to learn how to manage complexity
by building hierarchical systems of well-encapsulated, reusable components.
It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
as well as RSEs that can manage complex systems of almost black-box components,
and the future of RSE education should aim to prepare students for both roles.

@@

\subsection{Ethics and social considerations}
\begin{whatis}{}{ethics}
The principles of Good Scientific Practice \cite{dfg_gsp} have been a cornerstone in the conduct of research.
But their relevance hinges on constant updates in light of how research is performed.
By 2035 we want to have achieved accepted rules on the use of AI in research tasks and software development.
Open Science and Open Research are well-established in academia, and supported by respective Open Research Software
strategies of the political actors. We also hope that Data Awareness and Data Protection are still values
that are protected.
\end{whatis}

@@

% institutional bonds and ethics
Sometimes these values may conflict with the aims of their peers or institution. For example, their PI might insist on quickly developing some software without adhering to current best practices.
A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
This can then form the basis for institutional policies and guidelines.

% society implications
The work being undertaken by RSEs may well have direct, or indirect, ethical implications in areas such as social engineering [TODO find source], climate [TODO find source] and more.
Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
Like engineers, RSEs take the responsibility for the software they develop.
RSEs need to be aware of these issues and develop their own ``red lines''.

@@

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is now working in a project that develops
software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
at the university, and increasingly struggles to find compelling Open Source alternatives that enables them keep complete control of their data
rather than relying on software and tools that are backed by large corporate entities. So far she still has the support of the university board, since it values
strategic autonomy.
Her son Kyle on the other hand started out with a degree in Physics and then took one of the newly created RSE Master's courses,
but quickly decided that work in academia with its time limited contracts is not attractive.
Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
and therefore have deprioritized publicly funded research. Kyle decided to go to an AI hardware company since they offer
more secure employment in this polarized world.
\end{story}

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{}{academic}
We expect research software to be a central part of scientific practice on the same level as data and publications.
This will be supported by the first generations of RSE-Master graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, that are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects \cite{Kempf2025-draft}.
We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which have them, we want to see that researchers naturally accept them as part of their available services.
Additionally researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks. This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years such as
the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
But we could imagine new RSE specialisations emerging in the fields
of quantum computing, such as the Quantum-RSE, and in federated online platforms for research.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

On the other hand the existence of people with a certain specilised skillset opens up possiblities for pooling.
Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernisation, cleanup, and modularisation,
which can pay off in faster and easier development in the future.
This is an opportunity that would not be possible if relying on short-term funded solo RSEs.

\emph{Which challenges will still persist}\\
The workshop participants expect that there will still be budgeting and administrative hurdles to creating
permanent and well-funded RSE teams.
Here it can only be strongly hoped, that, with the recognized importance of research software, state-actors
support the development with a reliable stream of funding.
This would also require a change of culture in research about the acceptance of research software.\\
While we hope that the barrier to researchers diminishes, the workshop participants
still expect some work to do. We hope that the compartmentalisation that occurs will help researchers reprioritize their workload.
An upcoming challenge will be that, with the emergence of more dedicated RSE units,
networking between them will come into the foreground, because there should be a strong need to share best practices
among institutions with successful RSE Teams.
On a more organisational level, we expect the sharing of models on how to organise different RSEs with different areas of specialist expertise, to fit the different organisational constraints.
After that, organisations will have to come up with updated models for hiring people.
While a standardized RSE Masters makes it easier to develop the next generation of practitioners with the necessary technical skills, it will be interesting to see if
RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE Masters degree ensures a hiring process on a more standardised level.

\begin{story}{Changes to the Universities}{changes}
Kim found work years ago at one of the first RSE units in Arcadia. After a couple of years he moved to the university of
Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
Kim improves upon the model set forth by the University of Arcadia based on valuable experience. Since the university
of Avalon has to follow different rules, not everything can be straightforwardly adapted. Life would be so much easier
if at least university regulations would be homogenized, no?
\end{story}

\emph{New challenges}
If we follow the ideas laid out in \cite{Kempf2025-draft}, then in addition to university wide RSE units,
there will be embedded RSEs in the departments. Shaping and developing these network bonds will be a task for the future.
While networking embedded RSEs and RSE units is essential for building robust, sustainable research software infrastructure, the question arises whether every group truly needs to host and manage its own infrastructure.
When each team or institution hosts its own systems, this often leads to siloed environments where resources, expertise, and data are isolated, limiting collaboration and creating inefficiencies. Such silos can result in duplicated efforts, higher costs, and challenges in maintaining data integrity and governance, as each group independently manages similar infrastructure and workflows. While local control offers flexibility, it also increases complexity and makes it harder to share innovations or best practices across the wider research community. To overcome these challenges, federated infrastructure models are needed, enabling shared access, interoperability, and coordinated development while still allowing for some local autonomy \cite{Jelinek2021}.
Federated approaches break down silos, promote knowledge exchange, and ensure that infrastructure investments benefit the broader ecosystem rather than just isolated groups.
